Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression

### 2020-05-19 paper/appendix.tex
n -> m

Equation \ref{eq:model} does not explicitly contain a measurement noise term.
Yet, in most experimental cases, the problem is best described as:
\begin{equation}
  Y = (XS+N)F+M
  \label{eq:model_noisy_measure}
\end{equation}
  with $M\in \mathbb{R}^{n \times d_y}$.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

>>>>

Equation \ref{eq:model} does not explicitly contain a measurement noise term.
Yet, in most experimental cases, the problem is best described as:
\begin{equation}
  Y = (XS+N)F+M
  \label{eq:model_noisy_measure}
\end{equation}
  with $M\in \mathbb{R}^{m \times d_y}$.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_y}$.}

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{m \times d_x}$, $X_{test} \in
\mathbb{R}^{m' \times d_x}$, $Y_{train} \in \mathbb{R}^{m\times d_y}$, $Y_{test}
\in \mathbb{R}^{m'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x, d_z}$.} $H, G = \text{CCA}(X_{train},
Y_{train})$\; $R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

### 2020-05-19 paper/fmri_exps.tex
n -> m

The preprocessed volumetric fMRI data was linearly projected to its closest surface
using nilearn \texttt{vol\_to\_surf} function with a 3 mm radius. A surface searchlight
analysis was then implemented by concatenating the surface
fMRI data within a 8mm radius of each vertex. For each subject and vertex, we
thus build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 1400 words
by $d_y\approx40$ vertices per searchlight sphere. Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

@@

This procedure yields an $X \in \mathbb{R}^{n \times d_x}$ matrix of $n\approx$
840 TR (Repetition Time: 2 sec.) by
$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
have a zero-mean and a unit variance.

@@

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (n=90) using the average $\Delta R$ across time. Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.

>>>>

The preprocessed volumetric fMRI data was linearly projected to its closest surface
using nilearn \texttt{vol\_to\_surf} function with a 3 mm radius. A surface searchlight
analysis was then implemented by concatenating the surface
fMRI data within a 8mm radius of each vertex. For each subject and vertex, we
thus build an observation matrix $Y \in \mathbb{R}^{m \times d_y}$ of $m\approx$ 1400 words
by $d_y\approx40$ vertices per searchlight sphere. Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

@@

This procedure yields an $X \in \mathbb{R}^{m \times d_x}$ matrix of $m\approx$
840 TR (Repetition Time: 2 sec.) by
$d_x=4$ features for each subject. Each of the columns of $X$ is normalized to
have a zero-mean and a unit variance.

@@

Each model was implemented for each subject and each time sample independently. Pairwise comparison between models were performed using a two-sided Wilcoxon test across subjects (m=90) using the average $\Delta R$ across time. Corresponding effect sizes are shown in Figure~\ref{fig:meg_results}, and p-values are reported below.

### 2020-05-19 paper/meg_exps.tex
n -> m

The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we
used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
and 10Hz for the low and high cut-off frequencies. The raw data was then
segmented 100ms before word onset and 1s after
word onset ($t=0$ms corresponds to word onset). Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240Hz. The average responses across words is
displayed in Figure \ref{fig:meg_evoked}.
For each subject and each time sample relative to word onset, we
build an observation matrix $Y \in \mathbb{R}^{n \times d_y}$ of $n\approx$ 2,700 words
by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

We use the same features as for the fMRI experiments, except that the features
were not convolved by an hemodynamic response function. This yields an $X \in \mathbb{R}^{n \times d_x}$
matrix of $n\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
the columns of $X$ is normalized to have a mean and a standard deviation of 0
and 1 respectively.

>>>>

The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we
used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
and 10Hz for the low and high cut-off frequencies. The raw data was then
segmented 100ms before word onset and 1s after
word onset ($t=0$ms corresponds to word onset). Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240Hz. The average responses across words is
displayed in Figure \ref{fig:meg_evoked}.
For each subject and each time sample relative to word onset, we
build an observation matrix $Y \in \mathbb{R}^{m \times d_y}$ of $m\approx$ 2,700 words
by $d_y=301$ MEG channels (273 magnetometers and 28 compensation channels). Each
of the columns of $Y$ is normalized to have zero mean and unit variance.

We use the same features as for the fMRI experiments, except that the features
were not convolved by an hemodynamic response function. This yields an $X \in \mathbb{R}^{m \times d_x}$
matrix of $m\approx$ 2,700 words by $d_x=4$ features for each subject. Each of
the columns of $X$ is normalized to have a mean and a standard deviation of 0
and 1 respectively.

### 2020-05-19 paper/paper.tex
n -> m

  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{n \times d_y}$}};
  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annS) at (\posS, -1.8){\scalebox{0.85}{$S \in \mathbb{D}^{d_x \times d_x}$}};
  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{n \times d_x}$}};
  \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

@@

We consider the measurement of multivariate signal $Y \in \mathbb{R}^{n \times
d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{n \times
d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $S
\in \mathbb{D}^{d_x \times d_x}$, we denote by $XS$ the causal factors of $Y$.
%
In summary, the problem can be formalized as:
%
\begin{equation}
    y_i = (x_i S + n_i) F
    \label{eq:model}
\end{equation}
%
where $i$ is a given sample, and $n_i$ is a sample-specific noise drawn from a
centered distribution.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of
them to have any form of covariance.
% what is a general covariance matrix?
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This problem space, along with the sizes of all variables involved, is
illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
Regression (B2B) is to estimate the matrix $S$, i.e. to identify the factors
that most reliably account for the multivariate observations.

@@

% JR: shall we update algorithm with LOO trick?
\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{S} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{S} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{S} \leftarrow \hat{S} + \text{diag}(\hat{H})$\;
    }
    $\hat{S} \leftarrow \hat{S} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{S}, Y)$\;
    \Return{$\hat{S}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

In the presence of noise, the causal factors of $S$ correspond to the positive
elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
the Appendix (\ref{recovering}).

>>>>

  \node[] (annY) at (\posY, -1.8){\scalebox{0.85}{$Y \in \mathbb{R}^{m \times d_y}$}};
  \node[] (annX) at (\posX, -1.8){\scalebox{0.85}{$X \in \mathbb{R}^{m \times d_x}$}};
  \node[] (annS) at (\posS, -1.8){\scalebox{0.85}{$S \in \mathbb{D}^{d_x \times d_x}$}};
  \node[] (annN) at (\posN, -1.8){\scalebox{0.85}{$N \in \mathbb{R}^{m \times d_x}$}};
  \node[] (annF) at (\posF, -1.8){\scalebox{0.85}{$F \in \mathbb{R}^{d_x \times d_y}$}};

@@

We consider the measurement of multivariate signal $Y \in \mathbb{R}^{m \times
d_y}$ (the dependent variables), generated from a set of putative causes $X \in \mathbb{R}^{m \times
d_x}$ (the independent variables), via some unknown linear apparatus $F \in \mathbb{R}^{d_x \times d_y}$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $S
\in \mathbb{D}^{d_x \times d_x}$, we denote by $XS$ the causal factors of $Y$.
%
In summary, the problem can be formalized as:
%
\begin{equation}
    y_i = (x_i S + n_i) F
    \label{eq:model}
\end{equation}
%
where $i$ is a given sample, and $n_i$ is a sample-specific additive noise drawn from a
centered distribution.
%
While the triplet of variables $X$ and $N$ are independent, we allow each of
them to have any form of covariance.
% what is a general covariance matrix?
%
In practice, we observe $m$ samples $(X, Y)$ from the model.
%
This problem space, along with the sizes of all variables involved, is
illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\cref{eq:model}, the goal of Back-to-Back
Regression (B2B) is to estimate the matrix $S$, i.e. to identify the factors
that most reliably account for the multivariate observations.

@@

% JR: shall we update algorithm with LOO trick?
\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{m \times d_x}$, output data $Y \in \mathbb{R}^{m\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{S} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{S} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{S} \leftarrow \hat{S} + \text{diag}(\hat{H})$\;
    }
    $\hat{S} \leftarrow \hat{S} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{S}, Y)$\;
    \Return{$\hat{S}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

In the presence of additive noise, the causal factors of $S$ correspond to the positive
elements of $\text{diag}(\hat H)$. The methods to recover them are presented in
the Appendix (\ref{recovering}).

### 2020-05-19 paper/synthetic.tex
n -> m

The data generating process for each experiment constructs $n=1000$ training examples
according to the model $Y = (\text{h} XS + N)F$, where $\text{h}$ is a
scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{n \times d_x}$
%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{n
%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $S \in
%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
%entries drawn from $\mathcal{N}(0, \sigma^2)$, \item $\Sigma_N = BB^\top$,
%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
%$\mathcal{N}(0, \sigma^2)$, \end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
$X \in \mathbb{R}^{n \times d_x}$ contains rows
drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{n \times d_x}$
contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $S \in \mathbb{R}^{d_x
\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
\times d_x}$ contains entries drawn from $\mathcal{N}(0, \sigma^2)$, and the
factor $\text{h} \in \mathcal{R}_+$.

@@

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(S, \hat S)$.
%\begin{equation} AUC = 1 - \sum_1^n (S_k - S_{k-1}) ( \hat{S}_k +
%\hat{S}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup, but not in the neuroimaging experiments below.

>>>>

The data generating process for each experiment constructs $m=1000$ training examples
according to the model $Y = (\text{h} XS + N)F$, where $\text{h}$ is a
scalar that modulates the signal-to-noise ratio.
%
Here,
%\begin{itemize} \item $F \in \mathbb{R}^{d_x \times d_y}$ contains entries
%drawn from $\mathcal{N}(0, \sigma^2)$, \item $X \in \mathbb{R}^{m \times d_x}$
%contains rows drawn from $\mathcal{N}(0, \Sigma_X)$, \item $N \in \mathbb{R}^{m
%\times d_x}$ contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, \item $S \in
%\mathbb{R}^{d_x \times d_x}$ is a binary diagonal matrix containing $n_c$ ones,
%\item $\Sigma_X = AA^\top$, where $A \in \mathbb{R}^{d_x \times d_x}$ contains
%entries drawn from $\mathcal{N}(0, \sigma^2)$, \item $\Sigma_N = BB^\top$,
%where $B \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
%$\mathcal{N}(0, \sigma^2)$, \end{itemize}
    $F \in \mathbb{R}^{d_x \times d_y}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$ where $\sigma^2$ is inversely proportional to $d_x$,
$X \in \mathbb{R}^{m \times d_x}$ contains rows
drawn from $\mathcal{N}(0, \Sigma_X)$, $N \in \mathbb{R}^{m \times d_x}$
contains rows drawn from $\mathcal{N}(0, \Sigma_N)$, $S \in \mathbb{R}^{d_x
\times d_x}$ is a binary diagonal matrix containing $n_c$ ones, $\Sigma_X =
AA^\top$ where $A \in \mathbb{R}^{d_x \times d_x}$ contains entries drawn from
$\mathcal{N}(0, \sigma^2)$, $\Sigma_N = BB^\top$ where $B \in \mathbb{R}^{d_x
\times d_x}$ contains entries drawn from $\mathcal{N}(0, \sigma^2)$, and the
factor $\text{h} \in \mathcal{R}_+$.

@@

Finally, to estimate whether each model accurately identifies causal factors independently of
their potential biases, we compute the
area-under-the-curve (AUC) across factors $AUC(S, \hat S)$.
%\begin{equation} AUC = 1 - \sum_1^m (S_k - S_{k-1}) ( \hat{S}_k +
%\hat{S}_{k-1}) / 2 \end{equation}
By definition, this AUC evaluation can only be done when ground truth labels are available, as is the case in
this synthetic setup, but not in the neuroimaging experiments below.
